Climatic warming strengthens a positive feedback between alpine shrubs and fire

### 2016-03-08 manuscript/manuscript.Rnw
More edits

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment into arctic and alpine grasslands. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}) and that warming doubled tall shrub seedling growth rates and may increase survival. Warming had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increase survival.
\end{abstract}

@@

Field manipulative experiments have been widely used to investigate the effects of climate change on plant communities. These experiments have provided invaluable information on how climate change directly influences plant phenology\cite{Hoffmann:2010kf, Dorji:2013tt}, reproduction\cite{Klady:2011dw}, morphology\cite{Hudson:2009cn}, growth\cite{Hollister:2005vf}, floristic composition\cite{Elmendorf:2012kz} and biotic interactions\cite{Klanderud:2005uw}. However, because disturbances vary both spatially and temporally, most have focused on mature plant responses in relatively undisturbed vegetation\cite{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions (but see\cite{Munier:2010cg,Graae:2010ir,Camac:2015hja}). Consequently, little is known about how climate change influences seedling demographic rates\cite{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions. We consider these issues in an Australian alpine heathland.

@@

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine and tundra ecosystems, nor do we know whether rates of seedling growth and survival will be improved under warmer post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, and whether grasses have competitive, facilitative or no effect on shrub seedling demographic rates, or whether such effects are altered by warmer post-fire conditions.

@@

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape. Under global warming, this plant community is likely to encroach grasslands and is itself susceptible to shrub thickening\cite{McDougall:2003gb}. To determine how warmer and drier post-fire conditions affects shrub cover, we combined field observations with a warming experiment that used seedlings of four dominant Australian evergreen obligate seeding alpine shrubs \textemdash \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) and \textit{Phebalium squamulosum} (Rutaceae; a tall shrub). This approach allowed us to examine: 1) how warmer post-fire conditions influence rates of both grass recovery and shrub seedling recruitment, growth and mortality; 2) how proximity to grass affect shrub seedling demographic rates, and 3) whether such affects change under warmer conditions post-fire conditions.

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (two dominant shrubs of open heathland). In 2012-13, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a 200 m altitudinal range (equivalent to a 1.5\textdegree C change in mean temperature). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

Knowing shrub seedlings were more abundant in burnt vegetation, we investigated whether warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} emerging after a wildfire were transplanted into experimentally burnt plots that were either passively warmed by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C using open top chambers (OTCs) or subject to current ambient conditions (controls). After \Sexpr{n_days(chamber_dates, FALSE)} days \Sexpr{n_days(chamber_dates, TRUE)} growing season days where OTC treatment was applied) or \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} years growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium}) growing in post-fire bare ground were taller in warmed plots by \Sexpr{ht_diff(non_tussock_otc_growth_data,'Grevillea')}, \Sexpr{ht_diff(non_tussock_otc_growth_data,'Prostanthera')} and \Sexpr{ht_diff(non_tussock_otc_growth_data,'Phebalium')} cm respectively (Fig. \ref{fig3}). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no significant difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, mean annual predicted height growth rates of \textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium} were \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Grevillea')}, \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Prostanthera')} and \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Phebalium')} times that observed in control plots, respectively. Projecting these differences equates to an average 6 cm tall seedling (the mean initial size used in this experiment) growing in warmed conditions reaching maximum mature height in \Sexpr{hmax(non_tussock_otc_growth_model,'Grevillea')} fewer years for \textit{Grevillea} and \Sexpr{hmax(non_tussock_otc_growth_model,'Prostanthera')} fewer years for both \textit{Prostanthera} and \textit{Phebalium} (Fig. \ref{fig3}). \textit{Asterolasia} was predicted to reach a mature height in approximately \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[2])} to \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[1])} years regardless of warming treatment.

We validated the \textit{Grevillea} and \textit{Asterolasia} growth responses observed in experimentally warmed plots by using an elevation gradient in open heathland and maximum height data for 2003 post-fire recruits (i.e. suspected 9 year old seedlings). Here, a 200 m altitudinal range is equivalent to an ambient temperature difference of approximately \Sexpr{0.77 * (200/100)}\textdegree C\cite{Brown:1989vx}, which is comparable to that observed between experimental warmed and control plots (\Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C). In response to shifts in temperature, maximum heights of Grevillea post-fire recruits were expected to decrease with elevation, whereas \textit{Asterolasia} seedlings were not expected to show this pattern. Our experimental predictions were verified (Supplementary Fig. S4). Mean maximum heights of \textit{Grevillea} seedlings in burnt open heathland were \Sexpr{max_ht_alt(greaus_seedling_max_ht_model)} cm taller at 1670 m a.s.l compared to seedlings at 1860 m a.s.l. (\Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[1]} cm vs \Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[2]} cm; a difference comparable to our experimental findings). In contrast, mean maximum heights of \textit{Asterolasia} seedlings did not vary significantly with elevation. Topographic Wetness Index (TWI) and fire severity had no detectable influence on maximum seedling heights in either species.

@@

After five years and across all plots, \Sexpr{total_deaths(non_tussock_otc_mortality_data,TRUE)}\% (\Sexpr{total_deaths(non_tussock_otc_mortality_data,FALSE)} out of 511) of all seedlings growing in burnt bare ground patches had died. Most deaths occurred in \textit{Asterolasia} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Asterolasia')}) followed by \textit{Phebalium} (\Sexpr{death_counts(non_tussock_otc_mortality_data, 'Phebalium')}), \textit{Grevillea} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Grevillea')}) and \textit{Prostanthera} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Prostanthera')}). \textit{Prostanthera} showed the largest treatment effect (Fig. \ref{fig4}), with annual mortality rates estimated to be near \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','otc')}\% in warmed plots and \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','ctl')}\% in control plots. This significant increase in survival may be a consequence of OTCs reducing the severity of spring frosts by rising minimum ambient and soil temperatures by \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean_min','grand_mean')}\textdegree C and \Sexpr{micro_clim(overall_microclimate_summary,'CTemp_3cmBG','mean_min','grand_mean')}\textdegree C, respectively (see Supplementary Fig S5-7). Warming also reduced mean seedling mortality in \textit{Grevillea} and \textit{Phebalium} (Fig. \ref{fig4}); however, for both species, the effect was not statistically significant. By contrast, annual mortality rates in the prostrate shrub, \textit{Asterolasia}, were marginally higher in warmed plots, but again this effect was not significant (Fig. \ref{fig4}).

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems\cite{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and inter-tussock gap size (i.e. the proximity of shrub seedlings to grass tussocks) on the growth and survival of \textit{Grevillea} seedlings (Fig. \ref{fig5}).  As with the seedlings growing in bare ground, we detected a strong positive effect of warming on growth rates and a marginal, yet not significant, decrease in mortality. We also found no significant effect of inter-tussock gap radius on growth or mortality rates. However, for growth, we did detect a marginally significant interaction between warming and gap size, such that in warmed plots growth rates slightly increased with gap size. This suggests that warmer conditions may increase competition between shrubs and grasses, a finding consistent with the abiotic stress hypothesis\cite{Callaway:2002vb}, which suggests that temperature stress in alpine and tundra ecosystems could increase the strength of competitive interactions. However, this effect was small, and likely to have little biological consequence.

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still reduce shrub recruitment opportunities by infilling post-fire bare ground gaps at a faster rate. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses. However, the rate at which this occurred was not significantly influenced by a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C increase in temperature (Fig. \ref{fig6}).

@@

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig \ref{fig1}b). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems\cite{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates conditions that allow shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates, and potentially increase their survival. These demographic effects will ultimately result in shrub thickening and expansion into grasslands. These effects, coupled with field\cite{Williams:2006tl}, experimental\cite{Fraser:NHT5Kx_W} and paleoecological\cite{Higuera:2008dw} evidence indicate that shrubs are the most flammable vegetation component of alpine and tundra ecosystems, and suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine landscape under a warmer environment. This further strengthens the feedback\cite{Higuera:2008dw, Higuera:2009ga} by potentially increasing the frequency or severity of fires, which then create further recruitment opportunities for shrubs with little demographic impact of neighbouring tussock grass.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent obligate seeding shrubs reaching reproductive age and thereby exhausting the seedbank. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes including in Australia, where it would require current fire intervals of 50 to 100 years to decrease to less than 20 years \textemdash the time observed for the species in this study to reach reproductive maturity\cite{Williams:2008ba}. Furthermore, if reproductive output is related to plant size\cite{Wenk:2015ko}, then climatic warming may allow obligate seeding species to reach reproductive maturity sooner, and consequently, increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack\cite{Wipf:2009jh}, soil moisture\cite{Engler:2011vq} or herbivory\cite{Post:2008wz} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways\cite{MyersSmith:2011fp}.

@@

We investigated shrub dynamics in open heathland because it is a common and highly flammable plant community in the Australian Alps above 1600 m a.s.l\cite{Williams:2006tl}. It is also an ecotone between closed heathland (>70\% shrub cover) and tussock grassland consisting of \textit{Grevillea australis} shrubs interspersed among a \textit{Poa hiemata} sward. Aerial photography has revealed that plant community has acted as an \textit{invasion front} of shrub expansion into grassland. It is also experiencing shrub thickening by closed heathland dominants such as \textit{Prostanthera cuneata}.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5m$^2$) \textit{Grevillea australis} shrubs in open heathland to create patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance in open heathland burnt by wildfire. After creating the bare ground patches we collected seedlings of dominant alpine shrub species from a nearby (< 2 km) site of similar altitude burnt by a late 2006 wildfire. We collected shrub seedlings of two dominant open heathland species (\textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species (\textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland (\textit{Phebalium squamulosum} (Rutaceae; a tall shrub). All four species are obligate seeders post-fire (i.e. fire killed)\cite{Camac:2013eb}.

A total of 640 seedlings, 256 \textit{Grevillea} (half used in \textit{Poa} inter-tussock experiment \textemdash see below) and 128 for each of \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} were used. Four seedlings per species were randomly selected and transplanted into a 4\texttimes 4 square grid in the center of each burnt patch, with 14 cm between individuals and the edge of the patch dominated by resprouting tussock grass \textit{Poa hiemata}. To examine interactions between tussock grass and shrub seedlings we also randomly transplanted four additional \textit{Grevillea australis} into various sized inter-tussock gaps between burnt \textit{Poa hiemata} immediately surrounding the bare ground (Supplementary Fig. S6). The experimental site was fenced to prevent grazing by deer and horses. Rabbits or invertebrates also did not graze on these seedlings.

>>>>

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment in arctic and alpine grasslands. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}) and that warming doubled tall shrub seedling growth rates and can increase survival. Warming also had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing seedling survival.
\end{abstract}

@@

Field manipulative experiments have been widely used to investigate the effects of climate change on plant communities. These experiments have provided invaluable information on how climate change directly influences plant phenology\cite{Hoffmann:2010kf, Dorji:2013tt}, reproduction\cite{Klady:2011dw}, morphology\cite{Hudson:2009cn}, growth\cite{Hollister:2005vf}, floristic composition\cite{Elmendorf:2012kz} and biotic interactions\cite{Klanderud:2005uw}. However, because disturbances vary both spatially and temporally, most have focused on mature plant responses in relatively undisturbed vegetation\cite{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions (but see\cite{Munier:2010cg,Graae:2010ir,Camac:2015hja}). Consequently, little is known about how climate change influences seedling demographic rates\cite{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions. We consider these issues in Australian alpine heathland.

@@

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, and whether grasses have competitive, facilitative or no effect on shrub seedling demographic rates, or whether such effects are altered by warmer post-fire conditions.

@@

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape. Under global warming, this plant community is likely to encroach grasslands and is itself susceptible to shrub thickening\cite{McDougall:2003gb}. To determine how warmer post-fire conditions affects shrub cover, we combined field observations with a warming experiment that used seedlings of four dominant Australian evergreen obligate seeding alpine shrubs \textemdash \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub). This combination of data sources allowed us to examine: 1) landscape scale drivers of shrub seedling establishment; 2) how warmer post-fire conditions influence rates of both grass recovery and shrub seedling recruitment, growth and mortality; 3) how proximity to grass affect shrub seedling demographic rates, and 4) whether such grass effects change under warmer conditions post-fire conditions.

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of open heathland). In 2012-13, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a 200 m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

As shrub seedlings were more abundant in burnt vegetation, we investigated how warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Phebalium} and \textit{Prostanthera} emerging after a wildfire were transplanted into experimentally burnt plots that were either passively warmed by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C using open top chambers (OTCs) or subject to current ambient conditions (controls). After \Sexpr{n_days(chamber_dates, FALSE)} days (\Sexpr{n_days(chamber_dates, TRUE)} growing season days where OTC treatment was applied) or \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} years growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera}) growing in post-fire bare ground were taller in warmed plots by \Sexpr{ht_diff(non_tussock_otc_growth_data,'Grevillea')}, \Sexpr{ht_diff(non_tussock_otc_growth_data,'Phebalium')} and \Sexpr{ht_diff(non_tussock_otc_growth_data,'Prostanthera')} cm, respectively (Fig. \ref{fig3}). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no significant difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, mean annual predicted height growth rates of \textit{Grevillea}, \textit{Phebalium} and \textit{Prostanthera} were \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Grevillea')}, \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Phebalium')} and \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Prostanthera')} times that observed in control plots, respectively. Projecting these differences equates to an average 6 cm tall seedling (the mean initial size used in this experiment) growing in warmed conditions reaching maximum mature height in \Sexpr{hmax(non_tussock_otc_growth_model,'Grevillea')} fewer years for \textit{Grevillea} and \Sexpr{hmax(non_tussock_otc_growth_model,'Prostanthera')} fewer years for both \textit{Phebalium} and \textit{Prostanthera} (Fig. \ref{fig3}). \textit{Asterolasia} was predicted to reach a mature height in approximately \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[2])} to \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[1])} years.

We validated the \textit{Grevillea} and \textit{Asterolasia} growth responses observed in experimentally warmed plots by examining the effect of elevation on maximum height data collected from 2003 post-fire recruits across 40 open heathland sites. Here, a 200 m altitudinal range is equivalent to a mean ambient temperature difference of approximately \Sexpr{round(0.77 * (200/100),1)}\textdegree C\cite{Brown:1989vx}, which is comparable to that observed between experimental warmed and control plots (\Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C). In response to shifts in temperature, maximum heights of \textit{Grevillea} post-fire recruits were expected to decrease with elevation, whereas \textit{Asterolasia} seedlings were not expected to show this pattern. Our experimental predictions were verified (Supplementary Fig. S4). Mean maximum heights of \textit{Grevillea} seedlings in burnt open heathland were \Sexpr{max_ht_alt(greaus_seedling_max_ht_model)} cm taller at 1670 m a.s.l compared to seedlings at 1860 m a.s.l. (\Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[1]} cm vs \Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[2]} cm; a difference comparable to our experimental findings). In contrast, mean maximum heights of \textit{Asterolasia} seedlings did not vary significantly with elevation. Topographic Wetness Index (TWI) and fire severity had no detectable influence on maximum seedling heights in either species.

@@

After five years and across all plots, \Sexpr{total_deaths(non_tussock_otc_mortality_data,TRUE)}\% (\Sexpr{total_deaths(non_tussock_otc_mortality_data,FALSE)} out of 511) of all seedlings growing in burnt bare ground patches had died. Most deaths occurred in \textit{Asterolasia} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Asterolasia')}) followed by \textit{Phebalium} (\Sexpr{death_counts(non_tussock_otc_mortality_data, 'Phebalium')}), \textit{Grevillea} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Grevillea')}) and \textit{Prostanthera} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Prostanthera')}). \textit{Prostanthera} showed the largest treatment effect (Fig. \ref{fig4}), with annual mortality rates estimated to be near \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','otc')}\% in warmed plots and \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','ctl')}\% in control plots. This significant decrease in mortality may be a consequence of OTCs reducing the severity of spring frosts by rising minimum ambient and soil temperatures by \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean_min','grand_mean')}\textdegree C and \Sexpr{micro_clim(overall_microclimate_summary,'CTemp_3cmBG','mean_min','grand_mean')}\textdegree C, respectively (see Supplementary Fig S5-7). Warming also reduced mean seedling mortality in \textit{Grevillea} and \textit{Phebalium} (Fig. \ref{fig4}); however, for both species, the effect was not statistically significant. By contrast, annual mortality rates in the prostrate shrub, \textit{Asterolasia}, were marginally higher in warmed plots, but again this effect was not significant (Fig. \ref{fig4}).

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems\cite{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and inter-tussock gap size (i.e. the proximity of shrub seedlings to grass tussocks) on the growth and survival of \textit{Grevillea} seedlings (Fig. \ref{fig5}).  As with the seedlings growing with no nearby \textit{poa tussock}, we detected a strong positive effect of warming on growth rates and a marginal, yet not significant, decrease in mortality. We also found no significant effect of inter-tussock gap radius on growth or mortality rates. However, for growth, we did detect a marginally significant interaction between warming and gap size, such that in warmed plots growth rates slightly increased with gap size. This suggests that warmer conditions may increase competition between shrubs and grasses, a finding consistent with the abiotic stress hypothesis\cite{Callaway:2002vb}. However, this effect was small and unlikely to have significant biological consequence on shrub growth rates.

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still limit shrub recruitment, and thus, shrub expansion, by infilling post-fire bare ground gaps at a faster rate under warmer conditions. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses (Fig. \ref{fig6}). However, the rate at which this occurred was very slow, with a 10 cm radius gap decreasing by approximately \Sexpr{gap_size(gap_dynamic_model, 'otc')} cm over a ten year period. We also detected no significant effect of a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C temperature rise on the rate of grass infilling.

@@

Here we quantified several unknown interactions between fire, shrub-grass relationships and climate to extend a conceptual model of alpine shrub dynamics under climate change (Fig. \ref{fig1}b). We showed how these interactions strengthen a hypothesized feedback loop that can rapidly increase shrub cover in alpine and tundra ecosystems\cite{Higuera:2008dw}. The combination of rising temperatures and more frequent or severe fire creates conditions that allow shrub seedlings to establish in greater densities, and for tall shrubs, double their growth rates and potentially increase their survival. These demographic effects will ultimately result in shrub thickening and expansion into grasslands. These effects, coupled with field\cite{Williams:2006tl}, experimental\cite{Fraser:NHT5Kx_W} and paleoecological\cite{Higuera:2008dw} evidence indicate that shrubs are the most flammable vegetation component of alpine and tundra ecosystems, and suggest that flammable fuel loads will accumulate faster and cover a larger proportion of the alpine landscape under a warmer environment. This further strengthens the feedback\cite{Higuera:2008dw, Higuera:2009ga} by potentially increasing the frequency or severity of fires, which then create further recruitment opportunities for shrubs with little demographic impact of neighbouring tussock grass.

While we have addressed several unknowns associated with this feedback between climatic warming, shrubs and fire, there are others we have not addressed that may also strengthen, weaken or break this feedback. The most obvious mechanism that will break this cycle involves shorter fire intervals that prevent obligate seeding shrubs reaching reproductive age and thereby exhausting the seedbank. However, this scenario is unlikely for the majority of alpine (or tundra) landscapes including in Australia, where it would require current fire intervals of 50 to 100 years to decrease to less than 20 years \textemdash the time estimated for the species in this study to reach reproductive maturity\cite{Williams:2008ba}. Furthermore, if reproductive output is related to plant size\cite{Wenk:2015ko}, then climatic warming may allow obligate seeding species to reach reproductive maturity sooner, and consequently, increase species resilience to short fire intervals. Nevertheless, these factors and others such as changes in snow pack\cite{Wipf:2009jh}, soil moisture\cite{Engler:2011vq} or herbivory\cite{Post:2008wz} require further research because they are likely to be altered by the interactive effects of climate and disturbance in unpredictable ways\cite{MyersSmith:2011fp}.

@@

We investigated shrub dynamics in open heathland because it is a common and highly flammable plant community in the Australian Alps above 1600 m a.s.l\cite{Williams:2006tl}. It is also an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland consisting of \textit{Grevillea australis} shrubs interspersed among a \textit{Poa hiemata} sward. Aerial photography has revealed that the plant community has acted as an \textit{invasion front} of shrub expansion into grassland and has itself experienced shrub thickening by closed heathland dominants such as \textit{Prostanthera cuneata}.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5m$^2$) \textit{Grevillea australis} shrubs in open heathland to create patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance in open heathland burnt by wildfire. After creating the bare ground patches we collected seedlings of dominant alpine shrub species from a nearby (\textless 2 km) site of similar altitude burnt by a late 2006 wildfire. We collected shrub seedlings of two dominant open heathland species (\textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species (\textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland (\textit{Phebalium squamulosum} (Rutaceae; a tall shrub). All four species are obligate seeders post-fire (i.e. fire killed)\cite{Camac:2013eb}.

A total of 640 seedlings, 256 \textit{Grevillea} (half used in \textit{Poa} inter-tussock experiment \textemdash see below) and 128 for each of \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} were used. Four seedlings per species were randomly selected and transplanted into a 4\texttimes 4 square grid in the center of each burnt patch, with 14 cm between individuals and the edge of the patch dominated by resprouting tussock grass \textit{Poa hiemata}. To examine interactions between tussock grass and shrub seedlings we also randomly transplanted four additional \textit{Grevillea australis} into various sized inter-tussock gaps between burnt \textit{Poa hiemata} immediately surrounding the bare ground (Supplementary Fig. S8). The experimental site was fenced to prevent grazing by deer and horses. Rabbits or invertebrates also did not graze on these seedlings.

To simulate near-term warmer conditions predicted under IPCC\cite{IPCC:2013vv}, we randomly assigned Open Top Chambers (OTCs) to half (16) the plots, with the remainder treated as unwarmed controls. OTCs were placed over plots, ensuring all seedlings (including inter-tussock shrub seedlings) occurred within the 1.1 m$^2$ open top to minimise edge effects. OTCs were placed out at the start of the growing season (October) where they remained until snowfall (early June). This procedure was repeated for five growing seasons from May 2010 to May 2015.
